---
title: "Getting started with loan"
author: 'Rudolfs Kregers'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **loan** package analyses a lender's application funnel and loan book
together. This vignette introduces the core: how a dataset is declared, and how
every analysis function can be called.

```{r setup}
library(loan)
data(fintech)
```

## One row per closed application

`fintech` holds 183,532 applications of a digital consumer lender. Every row is
one **closed** application -- approved and turned into a loan, rejected, or
cancelled. Rejected and cancelled applications have no outcome, but they stay in
the data: approval rates and reject inference need them.

```{r}
table(fintech$application_status)
```

## Declaring the roles: `loan_tbl()`

`loan_tbl()` records which column plays which role, once. Afterwards the
analysis functions fill in their own arguments.

```{r}
lt <- loan_tbl(
  fintech,
  ## ids
  application_id         = "application_id",
  loan_id                = "loan_id",
  client_id              = "client_id",
  ## timestamps
  application_created_at = "app_created_at",
  ## application status, mapped onto the canonical approved / rejected / cancelled
  application_status     = value_map("application_status",
                                     approved  = "LOAN_ISSUED",
                                     rejected  = "REJECTED",
                                     cancelled = "CANCELLED"),
  ## outcomes: several may be declared, one is primary
  outcomes = list(
    fpd15           = binary_outcome("fpd15", bad = 1, good = 0),
    fpd31           = binary_outcome("fpd31", bad = 1, good = 0),
    defaulted       = binary_outcome("loan_status", bad = c("DEFAULTED", "WRITE_OFF")),
    dpd             = "current_dpd",
    extension_times = "extension_times",
    cash_in         = "cash_in"
  ),
  primary_outcome        = "fpd15",
  ## post-decision information: kept with the data, never analysed
  supplementary          = c("app_close_reason", "cash_out", "principal_disbursed")
)
lt
```

The roles are:

| Role | Holds |
|---|---|
| `application_id`, `loan_id`, `client_id` | identifiers |
| `application_created_at`, `first_delay_at` | timestamps |
| `application_status` | a column, or a `value_map()` onto `approved` / `rejected` / `cancelled` |
| `outcomes`, `primary_outcome` | one or several outcomes, and the default one |
| `predictors` | the features to analyse |
| `supplementary` | columns kept but never analysed |

Every column holds exactly one role; a column that is both an outcome and a
predictor is an error, because it would leak the outcome.

**Outcomes.** A binary outcome is declared with `binary_outcome()`, which says
which raw values are `bad`; everything is then counted as the canonical `good` /
`bad`. An outcome given by column name only (`dpd` above) has its type detected
when a function uses it.

**Unlisted columns.** Columns you do not list get a role from `unlisted_role`:
`"predictors"`, `"supplementary"`, or `"auto"` (the default) -- numeric,
logical and nominal columns become predictors, anything else supplementary. In
`"auto"` mode, a column that is a near-copy of one you listed (an id, a
timestamp, an outcome or a predictor; association above 0.95) becomes
supplementary instead. `loan_tbl()` reports what it assigned: review it, since
post-outcome columns that are not copies of anything would still become
predictors.

## Three ways to call an analysis function

Every analysis function accepts three input forms, chosen by its first argument.
They give the same result:

```{r}
fpd15_labels <- ifelse(fintech$fpd15 == 1, "bad", "good")
status <- value_map("application_status", approved = "LOAN_ISSUED",
                    rejected = "REJECTED", cancelled = "CANCELLED")

# 1. bare vectors
g1 <- group_stats(fintech$client_age, fpd15_labels,
                  application_status = fintech$application_status, status_map = status)

# 2. a data frame and column names
g2 <- group_stats(fintech, "client_age",
                  outcome = binary_outcome("fpd15", bad = 1),
                  application_status = status)

# 3. a loan_tbl: the roles fill in the rest
g3 <- group_stats(lt, "client_age")

identical(g1$approval_rate, g2$approval_rate) && identical(g2$approval_rate, g3$approval_rate)
identical(g1$bad_rate, g2$bad_rate) && identical(g2$bad_rate, g3$bad_rate)
```

Where a data frame is the first argument, a single string is a column name and
a full-length vector is the values themselves, so computed variables work too:
`group_stats(fintech, log1p(fintech$gross_salary), ...)`.

On a `loan_tbl`, the `outcome` argument chooses among the declared outcomes:

| You write | The outcome used |
|---|---|
| nothing | the `primary_outcome` |
| `outcome = "defaulted"` | the declared outcome of that name |
| `outcome = binary_outcome("loan_status", bad = "DEFAULTED")` | declared on the spot |

```{r}
group_stats(lt, "client_age", outcome = "defaulted")[, c("the_var", "bad_rate")]
```

A function that needs a binary outcome says so when given another type:

```{r, error = TRUE}
group_stats(lt, "client_age", outcome = "dpd")
```

The roles survive dplyr verbs, so a `loan_tbl` can be filtered and piped:

```{r}
lt |>
  dplyr::filter(loan_seq_number == 1) |>
  group_stats("client_age")
```

## Feature analysis

`contingency_table()` counts applications per group of a variable, by status
and by outcome. It is the building block of the other functions.

```{r}
contingency_table(lt, "gender")
```

`group_stats()` adds the approval rate -- cancelled applications never received
a risk decision, so they are excluded -- the bad rate, and optional statistics:
weight of evidence and a per-group Fisher exact test.

```{r, fig.width = 7, fig.height = 6}
gs <- group_stats(lt, "client_age", stats = c("woe", "fisher_p_val"))
gs
plot(gs)
```

Missing values are a group of their own: in `fintech`, applications without a
declared gender were never approved.

```{r}
group_stats(lt, "gender")[, c("the_var", "applications_total", "approval_rate", "bad_rate")]
```

`psi()` measures how much a variable's distribution moved between two samples,
typically two time periods. The comparison sample reuses the base sample's
groups, so both are binned identically.

```{r}
base <- as.Date(lt$app_created_at) <  as.Date("2024-04-01")
comp <- as.Date(lt$app_created_at) >= as.Date("2024-07-01")
psi(lt, "education_level", time_split_base = base, time_split_comparison = comp)
```

`plot_univariate_smooth()` draws a smoothed bad rate -- or approval rate --
across a numeric variable.

```{r, fig.width = 7, fig.height = 4}
plot_univariate_smooth(lt, "client_age")
```

## Status of the package

The data contract (`loan_tbl()`, `value_map()`, `binary_outcome()`) and the
analysis functions above are the stable core that further tools are built on.
`dynamic_stats()` (statistics over many time splits) and the modelling layer
(`evaluate_features()`, `fit_model()`, `cv_folds()`, `auto_recipe()`,
`step_woebin()`) are experimental and are being redesigned.
